Legion: Expressing Locality and Independence with Logical Regions

### 2012-05-04 doc/sc2012/experiments.tex

Figure \ref{fig:ckt_speed} shows the performance of the Legion circuit simulation relative
to a single-GPU implmementation hand-coded in CUDA.  Each line shows the scaling of
a particular problem size as the number of nodes is varied.  Our results demonstrate
excellent strong scaling behavior, with speedups of *??*x (compared to the single-GPU CUDA
version) for the small problem on 48 GPUs and *??*x for the larger problem size on 96 GPUs.

@@

Our second experiment is a port of the \emph{fluidanimate} benchmark from the PARSEC suite\cite{bienia11benchmarking},
which does a particle-based simulation of a fluid.  Each particle interacts only with other nearby
particles, so the benchmark divides the space in which the fluid
can move into a three-dimensional array of cells such that the range of interaction is limited to just the
cells adjacent (including diagonals) to the one a particle resides in.  Originally designed to be a
chip multiprocessor benchmark, the application divides the array into ``grids'' and assigns each one to 
a thread.  Per particle locking is used to safely access particles in cells that lie
on the edge of a grid.  This fine-grained locking along with the assumption of a shared address space lead to
good scaling in a multi-core processor, but rule out any attempt to scale beyond a single node.  To extend the
scaling further, our port divides uses region partitioning to divide the array into grids in a similar
way, but handles the interaction between grids in a way that doesn't rely on access to uniform shared memory.
Instead, the Legion version of the benchmark creates explicit ghost copies of cells on the boundaries of grids
and uses those ghost cells to exchange information between the grids.

>>>>

Figure \ref{fig:ckt_speed} shows the performance of the Legion circuit simulation relative
to a single-GPU implmementation hand-coded in CUDA.  Each line shows the scaling of
a particular problem size as the number of nodes is varied.  Our results demonstrate
excellent strong scaling behavior, with speedups of 39.0X (compared to the single-GPU CUDA
version) for the small problem on 48 GPUs and 62.5X for the larger problem size on 96 GPUs.

@@

Our second experiment is a port of the \emph{fluidanimate} benchmark from the PARSEC suite\cite{bienia11benchmarking},
which does a particle-based simulation of a fluid.  Each particle interacts only with other nearby
particles, so the benchmark divides the space in which the fluid
can move into a three-dimensional array of cells such that the range of interaction is limited to just the
cells adjacent (including diagonals) to the one a particle resides in.  Originally designed to be a
chip multiprocessor benchmark, the application divides the array into ``grids'' and assigns each one to 
a thread.  Per particle locking is used to safely access particles in cells that lie
on the edge of a grid.  This fine-grained locking along with the assumption of a shared address space lead to
good scaling in a multi-core processor, but rule out any attempt to scale beyond a single node.  To extend the
scaling further, our port uses region partitioning to divide the array into grids in a similar
way, but handles the interaction between grids in a way that doesn't rely on access to uniform shared memory.
Instead, the Legion version of the benchmark creates explicit ghost copies of cells on the boundaries of grids
and uses those ghost cells to exchange information between the grids.

### 2012-05-04 doc/sc2012/legion.tex

Modern parallel machines are increasingly complex, with deep,
distributed memory hierarchies and heterogeneous processing units.  Because the
costs of communication within these architectures vary by several orders of magnitude, 
the penalty for mistakes in the placement of data or computation is usually very poor performance.  Thus,
to achieve good performance it is crucial that the programmer
and the compiler be able to reason about {\em locality} (data resident
close to computation that uses it) and {\em independence} (computations
operating on disjoint data, and therefore not requiring communication and able to be placed in possibly distant parts of the machine).  
Most contemporary
programming systems have no facilities for the programmer to express locality
and independence.  The few languages that do focus primarily on
array-based locality \cite{Fatahalian06,CHAPEL04,UPC99} and 
avoid irregular pointer data structures.  
%
% FIXME
% Sean would like a bit more on why other systems don't support pointer data structures (e.g., the difficulty of
% alias analysis.  I thought about it and feel it is too much of a digression: mentioning that in the first paragraph
% makes it sound like it will be an important part of the paper.
%
%encode such structures in arrays, which 
%data structures encoded in arrays which are impossible for compilers
%and runtime systems to reason about with regard to locality and independence.

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data
and to make explicit relationships useful for reasoning about locality
and independence.
A logical region names a set of objects.  Logical regions are first-class 
values in Legion and may be dynamically allocated, deleted and stored
in data structures. Regions can also be passed as arguments
to distinguished functions called {\em tasks} that access the data in
those regions, providing locality information.  Logical regions may be
{\em partitioned} into (usually) disjoint {\em subregions}, providing
information for determining independence of computations.
Furthermore, computations access logical regions with particular {\em
  privileges} ({\em read-only}, {\em read-write}, and {\em reduce}) and {\em coherence} (e.g., {\em exclusive access} and {\em
  atomic access}, among others).  Privileges express how a task may use
its region arguments, providing data dependence information that
is used to guide the extraction of parallelism.  For example, if two tasks
access the same region with read-only privileges the two tasks
can potentially be run in parallel.
Coherence properties express the required semantics of concurrent region accesses.
For example, if the program executes $f_1(r); f_2(r)$ and tasks $f_1$
and $f_2$ both require {\em exclusive} access to region $r$, then Legion guarantees the result will
be as if $f_1(r)$ completes before $f_2(r)$ begins.   On the other hand, if
the tasks access $r$ with {\em atomic} coherence, then Legion
guarantees only atomicity of the tasks with respect to $r$: either task $f_1(r)$
appears to run entirely before $f_2(r)$ or vice versa.

Legion programs are written using logical regions, which do not commit
to any particular placement of the data in the machine.  At runtime,
each logical reagion has one or more {\em physical instances}.  Often
logical regions correspond to a single physical instance, but it is also common and useful
to have multiple physical copies of a logical region's contents in different parts of the machine 
to duplicate and distribute read-only data, to keep an old copy of a region to break a 
write-after-read dependence, or to allow independent, local reductions that are later combined.

@@

\item We present the results of experiments on three applications:
  fluid-flow on a regular grid, a three-level AMR code solving a
  standard heat diffusion equation and a circuit simulation.
  We run each of these four codes and best available reference
  versions on three different clusters of multicore processors with
  attached GPUs: a small cluster of our own (4 nodes with 12 CPUs and 2 GPUs per node), a medium-sized cluster
  designed for visualization applications (10 nodes with 12 CPUs and 5 GPUs per node), and up to a third of the
  Keeneland supercomputer (32 nodes, 12 CPUs and 3 GPUs per node).
%
% FIXME Give the same numbers as in the abstract.
%

>>>>

Modern parallel machines are increasingly complex, with deep,
distributed memory hierarchies and heterogeneous processing units.  Because the
costs of communication within these architectures vary by several orders of magnitude, 
the penalty for mistakes in the placement of data or computation is usually very poor performance.  Thus,
to achieve good performance it is crucial that the programmer
and the programming system be able to reason about {\em locality} (data resident
close to computation that uses it) and {\em independence} (computations
operating on disjoint data, and therefore not requiring communication and able to be placed in possibly distant parts of the machine).  
Most contemporary
programming systems have no facilities for the programmer to express locality
and independence.  The few languages that do focus primarily on
array-based locality \cite{Fatahalian06,CHAPEL04,UPC99} and 
avoid irregular pointer data structures.  
%
% FIXME
% Sean would like a bit more on why other systems don't support pointer data structures (e.g., the difficulty of
% alias analysis.  I thought about it and feel it is too much of a digression: mentioning that in the first paragraph
% makes it sound like it will be an important part of the paper.
%
%encode such structures in arrays, which 
%data structures encoded in arrays which are impossible for compilers
%and runtime systems to reason about with regard to locality and independence.

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data
and to make explicit relationships useful for reasoning about locality
and independence.
A logical region names a set of objects.  Logical regions are first-class 
values in Legion and may be dynamically allocated, deleted and stored
in data structures. Regions can also be passed as arguments
to distinguished functions called {\em tasks} that access the data in
those regions, providing locality information.  Logical regions may be
{\em partitioned} into disjoint or aliased {\em subregions}, providing
information for determining independence of computations.
Furthermore, computations access logical regions with particular {\em
  privileges} ({\em read-only}, {\em read-write}, and {\em reduce}) and {\em coherence} (e.g., {\em exclusive access} and {\em
  atomic access}, among others).  Privileges express how a task may use
its region arguments, providing data dependence information that
is used to guide the extraction of parallelism.  For example, if two tasks
access the same region with read-only privileges the two tasks
can potentially be run in parallel.
Coherence properties express the required semantics of concurrent region accesses.
For example, if the program executes $f_1(r); f_2(r)$ and tasks $f_1$
and $f_2$ both require {\em exclusive} access to region $r$, then Legion guarantees the result will
be as if $f_1(r)$ completes before $f_2(r)$ begins.   On the other hand, if
the tasks access $r$ with {\em atomic} coherence, then Legion
guarantees only atomicity of the tasks with respect to $r$: either task $f_1(r)$
appears to run entirely before $f_2(r)$ or vice versa.

Legion programs are written using logical regions, which do not commit
to any particular layout of the data or placement of the data in the machine.  At runtime,
each logical reagion has one or more {\em physical instances}.  Often
logical regions correspond to a single physical instance, but it is also common and useful
to have multiple physical copies of a logical region's contents in different parts of the machine 
to duplicate and distribute read-only data, to keep an old copy of a region to break a 
write-after-read dependence, or to allow independent, local reductions that are later combined.

@@

\item We present the results of experiments on three applications:
  fluid-flow on a regular grid, a three-level AMR code solving a
  standard heat diffusion equation and a circuit simulation.
  We run each of these four codes and best available reference
  versions on three different clusters of multicore processors with
  attached GPUs: a small cluster of our own (4 nodes with 12 CPUs and 2 GPUs per node), a medium-sized cluster
  designed for visualization applications (10 nodes with 12 CPUs and 5 GPUs per node), and up to a third of the
  Keeneland supercomputer (32 nodes, 12 CPUs and 3 GPUs per node) \cite{Keeneland}.
%
% FIXME Give the same numbers as in the abstract.
%

### 2012-05-04 doc/sc2012/mapping.tex

\subsection{Default Mapper}
\label{sec:defmapper}
To make writing Legion applications easier, we provide a default mapper implementation that
can be used to get an application working quickly at a moderate performance level.  In addition,
the default mapper serves as the base class for all custom mappers which makes implementing
custom mappers easier as they only have to override the functions that need to be specialized.

The default mapper employs a simple scheme for mapping tasks and regions.  When
{\tt select\_initial\_processor} is invoked for the default mapper, it first checks for which kind of processors
the task has variants (i.e. GPU).  If the fastest variant is for the kind of processor the mapper is managing
it will choose to keep the task local, otherwise it will send the task to the closest processor of the
fastest variant kind.

@@

For calls to {\tt map\_task\_region}, the default mapper constructs a stack of memories ordered from best-to-worst
by bandwidth from the local processor.  This stack is then returned as the location of memories to be used for
mapping each region.  Note that this is a greedy algorithm that works well for the common case, but can cause some 
regions to be pulled unnecessarily close to the processor and consuming precious fast memory space.

>>>>

\subsection{Default Mapper}
\label{sec:defmapper}
To make writing Legion applications easier, we provide a default mapper implementation that
can be used to get an application working quickly at a moderate performance level.  The 
default mapper employs a simple scheme for mapping tasks and regions.  When
{\tt select\_initial\_processor} is invoked for the default mapper, it first checks for which kind of processors
the task has variants (i.e. GPU).  If the fastest variant is for the kind of processor the mapper is managing
it will choose to keep the task local, otherwise it will send the task to the closest processor of the
fastest variant kind.

@@

For calls to {\tt map\_task\_region}, the default mapper constructs a stack of memories ordered from best-to-worst
by bandwidth from the local processor.  This stack is then returned as the location of memories to be used for
mapping each region.  Note that this is a greedy algorithm that works well for the common case, but can cause some 
regions to be pulled unnecessarily close to the processor and consume precious fast memory space.
